Ribosome accumulation during early phase resistance training in humans

### 2022-02-03 response-to-reviews-rev1.Rmd

---
title: "APH-2021-12-0652: Response to the reviewers"
author: 
  - Daniel Hammarström
  - Sjur J. Øfsteng
  - Nicolai B. Jacobsen
  - Krister B. Flobergseter
  - Bent R. Rønnestad
  - Stian Ellefsen
date: "2022-02-02"
output: pdf_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: references.bib
---

@@

> Could the authors please comment on the apparent disconnect between increase in RNA and RPS6 protein levels?

We thank the reviewer for raising this question as we have attempted to broaden our discussion. We interpret the results from our analysis as affected by a time-lag between increases in synthesis of total RNA (presumably rRNA) and RPS6. The model used "controls" for both quantities being increased over the whole training period, within a specific time-point there exists no relationship between RPS6 and total RNA. This separates our analysis from others who have shown correlations between RPS6 and rRNA without controlling for, in their case, degree of stimuli [@nakada2016] and without examining the time course of their increase. As we have already pointed to in the original manuscript, our results suggests different temporal regulation or turnover rates in total RNA and RPS6 protein levels. Since ribosomal proteins exists in excess (and are thus not rate limiting for ribosomal biogenesis) [@lam2007], it is tempting to assume constitutive expression is only indirectly related to the process of ribosomal biogensis. We have expanded the discussion by raising the point that unbound ribosomal proteins (including RPS6) may be passively increased by myonuclear accretion. Although somewhat speculative, this notion is supported by recent evidence suggesting a specific role of satellite cell fusion in increasing availability of ribosomal proteins together with increased rRNA transcription [@murach2022]. This could in part help explain our observations since an increased myonuclear pool is presumably maintained during short-term de-training but rRNA transcription is decreased (and/or ribophagy increased).

### 2022-02-03 manuscript.Rmd
updates

---
title: "Ribosome accumulation during early phase resistance training in humans"
author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  officedown::rdocx_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography:
  - resources/bibliography.bib
  - references.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

<!-- rpS6 dissconect from total RNA levels. -->
In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6. Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period. Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for. Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT. <span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam]. This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam]. Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined. However, a disconnect between RPS6 and total RNA after de-training could be explained by a greater number of myonuclei supporting a higher constitutive expression of ribosomal proteins. Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the myofibers ribosomal machinery [@murach2022]. However, the present study's observations are limited to a single ribosomal protein and lack data supporting RT-induced myonuclear accretion. A parallel mechanism that could help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis. 
[@RN2588].<!-- Warner 2009 General description of   -->  
</span>

>>>>

---
title: "Ribosome accumulation during early phase resistance training in humans"
author: Daniel Hammarström^1,2,£^&ast;, Sjur J. Øfsteng^2,£^&ast;, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  officedown::rdocx_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography:
  - resources/bibliography.bib
  - references.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

<!-- rpS6 dissconect from total RNA levels. -->
In the present study, training induced increases in rRNA and total RNA coincided with increases in rpS6. Changes in total RNA levels and rpS6 in response to de-training did however not correspond as rpS6 protein levels remained elevated after the de-training period. Training induced increases in rpS6 seen in the present study are in agreement to what has previously been reported in young men
[@RN1642],<!-- Brook 2015 rpS6 increase in resposne to RT over control -->
but not in elderly men and women where a decrease was observed in response to training despite increases in total RNA and rRNA
[@RN1755].<!-- Stec 2016 rpS6 decreased 22% after four weeks of training in older  -->
Although increases were seen in both rpS6 and and total RNA, rpS6 did not explain variations in total RNA when number of sessions were controlled for. Together with a disconnect after the de-training period, this suggest that regulations of rpS6 expression and ribosomal RNA transcription displays different temporal characteristics in response to RT. <span custom-style="CHANGE">Differences in temporal regulation between total RNA and ribosomal proteins may be explained by observations in cell culture indicating that excess amounts of ribosomal proteins are synthesized and imported into the nucleus [@lam2007]. This abundance assures the availability of ribosomal proteins for ribosome assembly upon rRNA transcription while also indicating that the availability of ribosomal proteins is not rate-limiting for ribosomal biogenesis [@lam2007]. Whether RT leads to an expanded pool of unbound ribosomal proteins, possibly through satellite cell expansion and myonuclear accretion in humans [@abousawan2021; @murach2022] remains to be determined. However, a disconnect between RPS6 and total RNA after de-training could be explained by a greater number of myonuclei supporting a higher constitutive expression of ribosomal proteins. Interestingly, in a recent study, Murach and colleagues showed that newly acquired myonuclei (from satellite cell fusion) contribute to the myofibers ribosomal machinery [@murach2022]. However, the present study's observations are limited to a single ribosomal protein and lack data supporting RT-induced myonuclear accretion. A parallel mechanism that could help explain our observations is the possibility of extra-ribosomal functions in selected ribosomal proteins affecting their expression independent of ribosomal biogenesis. 
[@RN2588].<!-- Warner 2009 General description of   -->  
</span>
